A pipeline to analyse time-course gene expression data

### 2019-05-15 scripts/manuscript.Rmd
Remove all namespace calls

- Add legends to the plots that don't have.
- Currently have two functions "plot_genes", "plot_centroids", that do exactly
  the same thing. Find a better name for both of them
- switch to camel case function names
- shorten function names
- Remove the namespace call
- Move the normalization to a supplementary data
- Make clear that it applies to RNASeq and how to call the different functions
  to RNASeq
- Check that clustering can deal with count data.

@@

**EAP: General comment. Your function names are incredibly long compared to
standard functions. Also, will want to get rid of `moanin::` in front of
functions. It's more precise, but an average user doesn't ever experience the
`::` command; usually just package creaters.**

@@

```{r}
variance_cutoff = 0.5
variance_per_genes = apply(data, 1, mad)
min_variance = stats::quantile(variance_per_genes, c(variance_cutoff))
variance_filtered_data = data[variance_per_genes > min_variance,]
```

```{r echo=FALSE, fig.width=6, fig.height=3, fig.caption="Distribution of variance"}
graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
	       ylab="# probes", main="")
graphics::abline(v=min_variance, col="#AB0000", lw=2)
```

@@

pca_data = stats::prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))
```

@@

```{r pca_plots, fig.width=6, fig.height=6}
graphics::par(mfrow=c(2, 2), mar=c(2.5, 2.5, 2.5, 2.5))
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")

graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC3")
graphics::plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC3")

@@

variance_filtered_data = variance_filtered_data[, ord]
data_corr = stats::cor(variance_filtered_data, method="pearson")
data_corr_meta = meta[ord, ]
data_corr_meta$Timepoint = as.factor(data_corr_meta$Timepoint)

NMF::aheatmap(
    data_corr,
    Colv=NA, Rowv=NA,
    annCol=data_corr_meta[, c("Group", "Timepoint")],
    annRow=data_corr_meta[, c("Group", "Timepoint")],
    annLegend=TRUE, 
    annColors=ann_colors,
    main="Correlation plot")
```

@@

If no formula is provided, it will default to the following: 
`formula = ~Group:splines::ns(Timepoint, df=degrees_of_fredoom) + Group + 0`

```{r}
splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
```

@@

```{r weekly_differential_analysis}
# Define contrast	
weekly_de_analysis =  moanin::DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)
```

@@

```{r echo=FALSE, results="hide"}
contrasts = create_timepoints_contrasts("M", "K", splines_model)
K_weekly_de_analysis =  moanin::DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "C", splines_model)
C_weekly_de_analysis =  moanin::DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "VH", splines_model)
VH_weekly_de_analysis =  moanin::DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

@@

contrasts = limma::makeContrasts(
	contrasts=timecourse_contrasts,
	levels=levels(meta$Group))

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
pvalues = moanin::DE_timecourse(
    data, splines_model, contrasts,
    use_voom_weights=FALSE)
qvalues = apply(pvalues, 2, stats::p.adjust)

@@

```{r}
log_fold_change_timepoints = moanin::estimate_log_fold_change(
    data, splines_model, contrasts,  method="timely")
```

@@

```{r}
log_fold_change_epicon = moanin::estimate_log_fold_change(
    data, splines_model, contrasts,  method="epicon")

log_fold_change_sum = moanin::estimate_log_fold_change(
    data, splines_model, contrasts,  method="sum")

log_fold_change_max = moanin::estimate_log_fold_change(
    data, splines_model, contrasts, method="max")

log_fold_change_min = moanin::estimate_log_fold_change(
    data, splines_model, contrasts, method="min")
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:12])
moanin::plot_genes(data[top_DE_genes_pval, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_epicon),
	 decreasing=TRUE)[1:12])
moanin::plot_genes(data[top_DE_genes_lfc, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)

@@

```{r filter_genes}
# Then rank by fisher's p-value and take max the number of genes of interest
# Filter out q-values for the pvalues table
fishers_pval = moanin::pvalues_fisher_method(pvalues)
fishers_qval = stats::p.adjust(fishers_pval)

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = moanin::splines_kmeans(
    y, splines_model, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
moanin::plot_centroids(kmeans_clusters$centroids, splines_model,
		       colors=ann_colors$Group,
		       smooth=TRUE)
```

@@

```{r}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = moanin::splines_kmeans_score_and_label(
    data, kmeans_clusters)
labels = scores_and_labels$labels

@@

both_labels = data.frame("kmeans"=kmeans_labels, "scored"=labels)
# Create another cluster.
#both_labels[is.na(both_labels)] = 21
confusion_matrix = table(both_labels)
text = as.matrix(as.character(confusion_matrix))
dim(text) = dim(confusion_matrix)
NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
	      Colv=NA)
labels = scores_and_labels$labels
labels = unlist(labels[!is.na(labels)])

@@

res = NMF::aheatmap(
    data_to_plot,
    Colv=NA,
    color="YlGnBu",
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=FALSE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(raw)"),
    treeheight=0, legend=FALSE)

# Now use the results of the previous call to aheatmap to reorder the genes.
NMF::aheatmap(
    moanin:::rescale_values(data_to_plot)[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)

@@

i = 1
for(n_cluster in all_possible_n_clusters){
    clustering_results = moanin::splines_kmeans(
	y, splines_model,
	n_clusters=n_cluster, random_seed=42,
	n_init=10)
    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
    all_clustering[[i]] = clustering_results$clusters
    i = i + 1
}
```

@@

```{r average_silhouette_method, fig.width=6, fig.height=4}
# function to compute average silhouette for k clusters
average_silhouette = function(labels, y) {
    silhouette_results = cluster::silhouette(unlist(labels[1]), dist(y))
    return(mean(silhouette_results[, 3]))
}

@@

```{r fig.width=6, fig.height=4}
stability_5 = read.table("results/stability_5.tsv", sep="\t")
consensus_matrix_stability_5 = moanin::consensus_matrix(stability_5,
							scale=FALSE)
NMF::aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
	      Colv=FALSE,
	      treeheight=0)

stability_20 = read.table("results/stability_20.tsv", sep="\t")
consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20,
							 scale=FALSE)
NMF::aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
	      Colv=FALSE,
	      treeheight=0)

@@

```{r fig.width=6, fig.height=6}
moanin::plot_model_explorer(all_labels)
```

@@

```{r fig.width=6, fig.height=6}
clusters = c("C6", "C7", "C8", "C9", "C10")
selected_labels = all_labels[clusters]
moanin::plot_model_explorer(selected_labels)
```

@@

```{r eval=FALSE}
moanin::plot_cdf_consensus(all_labels)
```

@@

```{r results="hide"}
ensembl = biomaRt::useMart("ensembl")
ensembl = biomaRt::useDataset("mmusculus_gene_ensembl", mart=ensembl)
```

@@

# convert gene names
genes = biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene"),
		filters="refseq_mrna", values=genes,
		mart=ensembl)["entrezgene"]
genes = as.vector(unlist(genes))
pathways = KEGGprofile::find_enriched_pathway(
    genes, species="mmu",
    download_latest=TRUE)$stastic

@@

```{r go_to_gene_mapping, results="hide", message=FALSE}
genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
	      values=gene_names,
	      filters="refseq_mrna",
	      mart=ensembl)

# Create gene to GO id mapping
gene_id_go_mapping = moanin::create_go_term_mapping(genes)
```

>>>>

- Add legends to the plots that don't have.
- Currently have two functions "plot_genes", "plot_centroids", that do exactly
  the same thing. Find a better name for both of them
- switch to camel case function names
- Remove the namespace call
- Make clear that it applies to RNASeq and how to call the different functions
  to RNASeq
- Check that clustering can deal with count data.

@@

@@

```{r}
variance_cutoff = 0.5
variance_per_genes = apply(data, 1, mad)
min_variance = quantile(variance_per_genes, c(variance_cutoff))
variance_filtered_data = data[variance_per_genes > min_variance,]
```

```{r echo=FALSE, fig.width=6, fig.height=3, fig.caption="Distribution of variance"}
hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
     ylab="# probes", main="")
abline(v=min_variance, col="#AB0000", lw=2)
```

@@

pca_data = prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))
```

@@

```{r pca_plots, fig.width=6, fig.height=6}
par(mfrow=c(2, 2), mar=c(2.5, 2.5, 2.5, 2.5))
plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")
plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC1")

plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC3")
plot(
    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    xlab="PC2", ylab="PC3")

@@

variance_filtered_data = variance_filtered_data[, ord]
data_corr = cor(variance_filtered_data, method="pearson")
data_corr_meta = meta[ord, ]
data_corr_meta$Timepoint = as.factor(data_corr_meta$Timepoint)

aheatmap(
    data_corr,
    Colv=NA, Rowv=NA,
    annCol=data_corr_meta[, c("Group", "Timepoint")],
    annRow=data_corr_meta[, c("Group", "Timepoint")],
    annLegend=TRUE, 
    annColors=ann_colors,
    main="Correlation plot")
```

@@

If no formula is provided, it will default to the following: 
`formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`

```{r}
splines_model = create_splines_model(meta, degrees_of_freedom=6)
```

@@

```{r weekly_differential_analysis}
# Define contrast	
weekly_de_analysis = DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)
```

@@

```{r echo=FALSE, results="hide"}
contrasts = create_timepoints_contrasts("M", "K", splines_model)
K_weekly_de_analysis =  DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "C", splines_model)
C_weekly_de_analysis =  DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "VH", splines_model)
VH_weekly_de_analysis =  DE_timepoints(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)

@@

contrasts = makeContrasts(
	contrasts=timecourse_contrasts,
	levels=levels(meta$Group))

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
pvalues = DE_timecourse(
    data, splines_model, contrasts,
    use_voom_weights=FALSE)
qvalues = apply(pvalues, 2, p.adjust)

@@

```{r}
log_fold_change_timepoints = estimate_log_fold_change(
    data, splines_model, contrasts,  method="timely")
```

@@

```{r}
log_fold_change_epicon = estimate_log_fold_change(
    data, splines_model, contrasts,  method="epicon")

log_fold_change_sum = estimate_log_fold_change(
    data, splines_model, contrasts,  method="sum")

log_fold_change_max = estimate_log_fold_change(
    data, splines_model, contrasts, method="max")

log_fold_change_min = estimate_log_fold_change(
    data, splines_model, contrasts, method="min")
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:12])
plot_genes(data[top_DE_genes_pval, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_epicon),
	 decreasing=TRUE)[1:12])
plot_genes(data[top_DE_genes_lfc, ], splines_model,
		   colors=ann_colors$Group, smooth=TRUE)

@@

```{r filter_genes}
# Then rank by fisher's p-value and take max the number of genes of interest
# Filter out q-values for the pvalues table
fishers_pval = pvalues_fisher_method(pvalues)
fishers_qval = p.adjust(fishers_pval)

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(
    y, splines_model, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
plot_centroids(kmeans_clusters$centroids, splines_model,
		       colors=ann_colors$Group,
		       smooth=TRUE)
```

@@

```{r}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = splines_kmeans_score_and_label(
    data, kmeans_clusters)
labels = scores_and_labels$labels

@@

both_labels = data.frame("kmeans"=kmeans_labels, "scored"=labels)
# Create another cluster.
#both_labels[is.na(both_labels)] = 21
confusion_matrix = table(both_labels)
text = as.matrix(as.character(confusion_matrix))
dim(text) = dim(confusion_matrix)
aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
	      Colv=NA)
labels = scores_and_labels$labels
labels = unlist(labels[!is.na(labels)])

@@

res = aheatmap(
    data_to_plot,
    Colv=NA,
    color="YlGnBu",
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=FALSE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(raw)"),
    treeheight=0, legend=FALSE)

# Now use the results of the previous call to aheatmap to reorder the genes.
aheatmap(
    moanin:::rescale_values(data_to_plot)[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)

@@

i = 1
for(n_cluster in all_possible_n_clusters){
    clustering_results = splines_kmeans(
	y, splines_model,
	n_clusters=n_cluster, random_seed=42,
	n_init=10)
    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
    all_clustering[[i]] = clustering_results$clusters
    i = i + 1
}
```

@@

```{r average_silhouette_method, fig.width=6, fig.height=4}
# function to compute average silhouette for k clusters
average_silhouette = function(labels, y) {
    silhouette_results = silhouette(unlist(labels[1]), dist(y))
    return(mean(silhouette_results[, 3]))
}

@@

```{r fig.width=6, fig.height=4}
stability_5 = read.table("results/stability_5.tsv", sep="\t")
consensus_matrix_stability_5 = consensus_matrix(stability_5,
							scale=FALSE)
aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
	 Colv=FALSE,
	 treeheight=0)

stability_20 = read.table("results/stability_20.tsv", sep="\t")
consensus_matrix_stability_20 = consensus_matrix(stability_20,
						 scale=FALSE)
aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
	 Colv=FALSE,
	 treeheight=0)

@@

```{r fig.width=6, fig.height=6}
plot_model_explorer(all_labels)
```

@@

```{r fig.width=6, fig.height=6}
clusters = c("C6", "C7", "C8", "C9", "C10")
selected_labels = all_labels[clusters]
plot_model_explorer(selected_labels)
```

@@

```{r eval=FALSE}
plot_cdf_consensus(all_labels)
```

@@

```{r results="hide"}
ensembl = useMart("ensembl")
ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
```

@@

# convert gene names
genes = getBM(attributes=c("ensembl_gene_id", "entrezgene"),
		filters="refseq_mrna", values=genes,
		mart=ensembl)["entrezgene"]
genes = as.vector(unlist(genes))
pathways = find_enriched_pathway(
    genes, species="mmu",
    download_latest=TRUE)$stastic

@@

```{r go_to_gene_mapping, results="hide", message=FALSE}
genes = getBM(attributes=c("go_id", "refseq_mrna"),
	      values=gene_names,
	      filters="refseq_mrna",
	      mart=ensembl)

# Create gene to GO id mapping
gene_id_go_mapping = create_go_term_mapping(genes)
```
